---
title: "Wrangle data from the NINA freshwater gillnett base"
author: "Kim M"
date: "2024-03-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set wd, load data, choose focal species and year of interest

The data contains gillnet catch data for multiple freshwater fish species, for our purpose in Freshrestore the most relevant species are brown trout (Salmo trutta, Art_formID 5) charr (Salvelinus alpinus, Art_formID 6). 


```{r species}
#load libraries
library(RFishBC)
library(FSA)
library(tidyverse)

setwd("~/FreshRestore/")
fish_data<-read.table("innlandsfisk.txt", header = TRUE, sep = ";", dec = ".")
focal_speciesid <- 5
#start_year <- 1967
#end_year <- 2017
#focal_lakes <-
```
#Incklude biotic variables
Include information whether a fish population is sympatric or allopatric, and also indicate if large predators are present (i.e., pike)
```{r species2}
allopatric_species <- fish_data %>%
  group_by(FK_InnsjoNr) %>%
  summarise(num_species = n_distinct(FK_ArtID)) %>%
  filter(num_species == 1) %>%
  select(FK_InnsjoNr)

fish_data$is_allopatric <- ifelse(fish_data$FK_InnsjoNr %in% allopatric_species$FK_InnsjoNr, "allopatric", "sympatric")

# add pike info
fish_data<- fish_data %>%
  group_by(FK_InnsjoNr) %>%
mutate(pike_presence = if_else("37" %in% unique(FK_ArtID), "Yes", "No")) #37 is pike

```


##Wrangling of data
Mostly cleaning and removing some of the data that looks strange. Also making ready for the RFishBC-function

```{r wrangling}
# subset data for focal species

species_data<-fish_data[fish_data$FK_ArtID==focal_speciesid,]


#remove NAs for Length and Radius 
backcalc_data<-species_data[complete.cases(species_data$Radius), ]
backcalc_data<-backcalc_data[complete.cases(backcalc_data$Lengde), ]
#Check fish data for different lakes
table(backcalc_data$FK_InnsjoNr)
# Some lakes have very few individuals across time(less than 10), remove these lakes as they dont seem representative for the species
# Remove rows with less than min_rows_per_category per category
backcalc_data <- backcalc_data %>%
  group_by(FK_InnsjoNr) %>%
  filter(n() >= 9) %>%
  ungroup()



plot(backcalc_data$Lengde~backcalc_data$Radius)
#Seem to be two types of measurements, just remove the "outliers" for now
backcalc_data<-backcalc_data[backcalc_data$Radius<400,]
#Cheking for strange length Radius correlations among lakes and years
data_check <- backcalc_data %>%
  group_by(FK_InnsjoNr, Aar) %>%
  summarize(R_squared = summary(lm(Lengde ~ Radius))$r.squared,
            Intercept = coef(lm(Lengde ~ Radius))[1])
print(data_check)

#decide on cut-off on 0.8
ok_lakes_years<-data_check[data_check$R_squared>0.8,]
#Remove r_squared of 1, which seems wronge
ok_lakes_years<-ok_lakes_years[!ok_lakes_years$R_squared==1,]

backcalc_data<-backcalc_data[backcalc_data$FK_InnsjoNr %in% ok_lakes_years$FK_InnsjoNr & backcalc_data$Aar %in% ok_lakes_years$Aar ,]
#plot(backcalc_data$Lengde~backcalc_data$Radius)

#setting up data for RFishBC (renaming variables)
backcalc_data <- backcalc_data %>%
  rename(id = InnlandsfiskID,
         agecap = AlderSkjell,
         radcap = Radius,
         rad1 = Radius1,
         rad2 = Radius2,
         rad3 = Radius3,
         rad4 = Radius4,
         rad5 = Radius5,
         rad6 = Radius6,
         rad7 = Radius7,
         rad8 = Radius8,
         rad9 = Radius9,
         rad10 = Radius10,
         rad11 = Radius11,
         rad12 = Radius12,
         rad13 = Radius13,
         rad14 = Radius14,
         rad15 = Radius15,
         rad16 = Radius16,
         rad17 = Radius17,
         rad18 = Radius18,
         rad19 = Radius19,
         rad20 = Radius20)





```

## Back calculation of length for individuals

Back calculation of lengths at age are based on scale radius

```{r backcalculation}

##Back calculating data using fraser lee per lake 
backcalc_data_FL <- backcalc_data %>%
  group_by(FK_InnsjoNr) %>%
  do(backCalc(., Lengde, BCM = "FRALE",
              inFormat = "wide", outFormat = "long", digits = 0)) %>%
  ungroup()


# Print the resulting dataframe
print(backcalc_data_FL)

###
#Rename variables
backcalc_data_FL <- backcalc_data_FL %>%
  rename(LengdeValder = bclen,
         alder_aar = ann)

#including year of length increment (specific year affecting specific growth)
backcalc_data_FL$vekstaar<-backcalc_data_FL$Aar-((backcalc_data_FL$agecap-backcalc_data_FL$alder_aar+1))
```

##Include climate variables
Including climate variables, idealy these would vary with year of growth

```{r climate variables}
#read climate var
library(readr)
expanded_climate_vars <- read_csv("expanded_climate_vars.csv")
names(expanded_climate_vars)

expanded_climate_vars$mean_summer<-rowMeans(expanded_climate_vars[c("may_temp_mean","jun_temp_mean","jul_temp_mean","aug_temp_mean")])

expanded_climate_vars$snow_des_mar<-rowSums(expanded_climate_vars[c("dec_snow_depth","mar_snow_depth")])

##match based on lakeID and year
#mean summer temp
backcalc_data_FL$mean_summer_temp <- expanded_climate_vars$mean_summer[match(paste(backcalc_data_FL$FK_InnsjoNr,backcalc_data_FL$vekstaar), paste(expanded_climate_vars$innsjo_nr,expanded_climate_vars$year))]

#Snow in spring and autumn 
backcalc_data_FL$snow_des_mar <- expanded_climate_vars$snow_des_mar[match(paste(backcalc_data_FL$FK_InnsjoNr,backcalc_data_FL$vekstaar), paste(expanded_climate_vars$innsjo_nr,expanded_climate_vars$year))]



```

