---
title: "Documentation"
author: "Kwaku Peprah Adjei"
bibliography: references.bib
format: 
  pdf:
    number-sections: true
    toc: false
    keep-tex: true
    latex_engine: xelatex
    cite-method: natbib
    colorlinks: true
    linkcolor: 'black'
    citecolor: "black"
    urlcolor: "black"
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
editor: visual
csl: methods-in-ecology-and-evolution.csl
output: 
  pdf_document:
    fig_crop: true
    keep_tex: true
    number-sections: true
    latex_engine: xelatex
header-includes: 
  - \usepackage{mathtools}
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage[a4paper, total={6in, 10in}]{geometry}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
  - \usepackage{amsmath,amssymb,amsfonts,amsthm}
  - \usepackage{multirow}
  - \usepackage{setspace}\doublespacing
  - \renewcommand{\abstractname}{Summary}
  - \usepackage{bm}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{rotating}
  - \usepackage{array}
  - \usepackage{doi}
  - \usepackage[sort, round]{natbib}
  - \usepackage{tikz}
  - \usepackage{float}
  - \usepackage{comment}
  - \usetikzlibrary{positioning}
  - \usetikzlibrary{shapes,decorations,arrows,calc,arrows.meta,fit,positioning}
indent: true
editor_options: 
  chunk_output_type: console
---

This document provides a walk-through on the model fitted for the freshRestore project. The document begins with the objectives for the modelling, dataset and covariates used and some results obtained.


# Objectives

-   Fit a population dynamic model (IPMs) with anthropogenic drivers \citep[human driven factors such as climate change, direct exploitation, pollution, biological invasions, sea-use change; ][]{moullec2021identifying}.

-   The IPM should be a mechanistic model \citep[plug and play;][]{frost2023integrated, smallegange2017mechanistic}. The IPMs should describe how the variables affect population.

-   IPMs should model temporal trends in human characteristics using variables that indicate catchment state of each year and a continuous time series data.

-   Produce prediction maps (of what?)

# Example dataset received

I have received three files which are located [here](https://github.com/Peprah94/fishyIPMs/tree/main/dataset) in the [gitHub repository](https://github.com/Peprah94/fishyIPMs/tree/main): 

 + *Example1.csv* : This file contains fish population data across lakes in Norway. The description of the column names in this file can be found [here](https://github.com/Peprah94/fishyIPMs/blob/main/dataset/Variable%20description.docx).
 
 + *Example2_vatnLnr.csv* : This file contains a individual fish data of the population harvested. The description of the column names in this file can be found [here](https://github.com/Peprah94/fishyIPMs/blob/main/dataset/Variable%20description.docx).
 
 + *model_catchment_vars.csv* : This file contains the catchment variables for all the lakes. The description of the variables can be found [here](https://github.com/Peprah94/fishyIPMs/blob/main/meetingScripts/Description%20of%20catchment%20variables%20.pdf).

# Data format and exploration

The individual fish data was formatted into an age at harvest format with their **lake**, **sex** and **year growth occurred** as variables to widen the long-format dataset received.

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(nimble)
library(dplyr)
library(readr)
library(reshape2)

load("dataset/formattedDataList.RData")
```


## Catchment Variables

-   Easily accessible variables e.g. using remote censored data

-   environmental covariates

I first perform a simple exploration of the catchment variables using *GGally* package. This is to help me reduce the number of catchment variables to include in the model. But since the end product expected is a plug and play type, I would have to include these catchment variables of interest even if they are not selected in the pre-screening phase.

```{r, echo=TRUE, warning=FALSE, eval=FALSE, message=FALSE}
# Not run
# copy and run if needed
GGally::ggpairs(dataList$ageAtHarvestData[, c(53:68)])
```

The following catchment variables are selected to model the fish survival in the lakes:

+ Coniferous forest

+ Moors and heathland

+ Peat bogs

+ Water bodies

+ Broad-leaved forest

+ Sparsely vegetated areas

+ mean_ndvi

## Other covariates

The **length at age this year** of the individual fishes was modelled with the following covariates:

+ Mean temperature in June

+ Precipitation in May

+ Precipitation in June

+ Precipitation in July

```{r, echo = TRUE, message=TRUE, eval=FALSE, warning=FALSE}
# Not run
# copy and run if needed
GGally::ggpairs(ageAtHarvestData[, c(1:3, 15, 36:45)])
```


# Fitted model

I fit a Bayesian age-at-harvest population model following \cite {skelly2023flexible, baerum2021population}. The script to run for the analysis can be found [here](https://github.com/Peprah94/fishyIPMs/blob/main/ModelFitting/modelFit.R). 

Here, I present a brief overview of the model fitted.

Let:

+ $i$ = individual

+ $a$ = age under consideration. Here the age is from $0$ to $10$.

+ $L_i$ = length at age this year for individual $i$

+ $S_{ia}$ = sprawning of individual $i$ at age $a$

+ $F_{ia}$ = Fecundicity of individual $i$ at age $a$


## Modelling $L_i$

$$
\begin{split}
L_i &\sim N(\lambda_L, \sigma^{2}_{L}) \\
\lambda_L &= X_L^T \beta_L
\end{split}
$$
where $X_L$ is a matrix of the covariates: Mean temperature in June, Precipitation in May, Precipitation in June and Precipitation in July; and $\beta_L$ is a vector of covariate effect plus an intercept term and $\sigma^{2}_{L}$ is the variation of the error term. 

The prior distributions assumed for $\beta_L$ and $\sigma^{2}_{L}$ are $N(0, 10^2)$ and $Uniform(0.1, 10)$ respectively.


## Sprawing probability

$$
\begin{split}
S_{ia} &\sim Bernouli(\psi_{ia}) \\
logit(\psi_{ia}) &= X_{sp}^T \beta_{sp} + \Gamma_a + \xi_{sex[i]}\\
\Gamma_a &\sim N(0, \sigma_a^2)\\
\xi_{sex[i]} &\sim N(0, \sigma_{sex}^2)
\end{split}
$$
where $X_{sp}$ is a matrix of covariates: age at year, $L_i$, age at year *  $L_i$, capture per unit effort; and $\beta_{sp}$ is a vector of covariate effect plus an intercept term. 


## Fecundicity
The equation for the fecundicity is derived from \cite{baerum2021population}.

$$
F_{ia} = exp(log(L_i * 2.21 - 6.15) \times S_{ia})
$$


## Survival Probability

$$
\begin{split}
logit(\gamma_{ia}) &= X_{surv}^T \beta_{surv} + \nu_a + \Xi_{lake[i]}\\
\nu_a &\sim N(0, \sigma_a^2)\\
\xi_{lake[i]} &\sim N(0, \sigma_{lake}^2)
\end{split}
$$
where $X_{surv}$ is a matrix of covariates described in section 3.1; and $\beta_{surv}$ is a vector of covariate effect plus an intercept term. 

## Projection Matrix (A)

The projection matrix used is the same as described in \cite{baerum2021population}. The reader is referred to page 9 of the [paper](https://www.nature.com/articles/s41598-021-94350-x).


## Model for observed counts at each age

The age at harvest count is modeled using the model described in \citep{skelly2023flexible}. The change we make is to estimate the stable population $\lambda$ from the largest eigenvalue of the projection matrix (A) instead of simulating it from an informed prior.

# Results

I am currently running the model with $20000$ iteration for each of the $2$ MCMC chains. It seems to take a while.

